---
title: "ST558 Homework 5"
format: html
editor: visual
---

```{r}
#| echo: false
#| message: true
library(tidyverse)
```

## Task 1: Read in the Data and Modify

We are using data from the [UCI machine learning repository](https://archive.ics.uci.edu/dataset/320/student+performance). The data is about secondary education in two Portuguese schools. First, I used the code they provided with the download to read in the data from a local file source and combine the two data frames.

```{r}

#Using the code given from the download package
d1=read.table("student-mat.csv",sep=";",header=TRUE)
d2=read.table("student-por.csv",sep=";",header=TRUE)

d3=merge(d1,d2,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet"))
print(nrow(d3)) # 382 students

```

Then I read in and combined the data using functions from the `tidyverse`. I used inner_join() on the variables they used in their code.

```{r}

d3=inner_join(d1,d2,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet"))

```

I then used an inner_join() on all variables other than G1, G2, G3, paid, and absences. I used this dataframe of the combined data in further exercises.

```{r}

#G1, G2, G3, paid, absences
d1_selected <- d1 %>% select(-G1, -G2, -G3, -paid, -absences)
d2_selected <- d2 %>% select(-G1, -G2, -G3, -paid, -absences)
d3_selected=inner_join(d1_selected,d2_selected,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet"))

```
